Mechanisation of Session-based Lambda Calculus Type Systems

### 2014-11-26 proposal/proposal.tex
Added discussion of Alms/Adoption&Focus/moved things around.

\usepackage{listings}
\usepackage{lstcoq}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{cmll}
\usepackage[style=alphabetic,natbib=true,maxbibnames=100]{biblatex}
\addbibresource{proposal.bib}

@@

The aim of this project is to develop a framework for formalising lambda
calculus type systems involving session types using interactive theorem
proving (in particular, the system Coq). The motivation for the framework is
that it will enable a whole class of related languages and type systems to be
formalised with little extra proof effort. Inspired by previous work on a
mechanised framework for $\pi$-calculus type systems~\cite{Gay:2001:FFP}, we
shift to the lambda calculus to provide a basis on which session types for
mainstream programming languages can be studied, whilst taking advantage of
previous proof effort and specialising only for those parts unique to the
particular system. The framework will be heavily based on calculi developed by \citeauthor{Wadler:2012}~\cite{Wadler:2012} describing a relationship between classical linear logic (CP) and a functional language with session types (GV). GV is a minimal functional language supporting linearity so I believe such a basis will be general enough to support reasoning about a large class of functional languages with similar base properties.

From this framework, a number of directions can be explored including
session polymorphism, session subtyping, and permitting different forms of
aliasing within the type system. The recent work by \citeauthor{Lindley:2014:SAP} \cite{Lindley:2014:SAP}, who extend GV to HGV supporting session forwarding, replication and polymorphism, would be an interesting place to start when looking to add these extensions.

Additionally, the majority of previous session type systems for mainstream programming languages have focussed exclusively on linear type systems. One disadvantage of a linear type system is the need to re-bind objects after each operation~\cite{Gay:2010:LAST}. One could extend the GV type system to provide the benefits of aliasing with the guarantees of linearity, termed ``adoption and focus'' as introduced by~\citeauthor{Fahndrich:2002} \cite{Fahndrich:2002}. The ``adoption and focus'' type system is non-trivial requiring special language features, e.g. to manage capabilities which can be thought as a form of object liveness mechanism, so the GV typing judgements must be extended such that the capabilities can be included in soundness results. A translation to CP must also be defined for the ``adopt'' and ``focus'' constructs.

@@

@@

Process reduction steps are equated with proof reduction via cut elimination. The relationship with intuitionistic linear logic provides absence of deadlocks in session-based communication for the process calculus.

@@

\citeauthor{Gay:2008:STI}~\cite{Gay:2008:STI} provide a $\lambda$-calculus based language with session types for channel primitives. The type system is nonstandard, featuring an additional environment for tracking the session types of channels; these types are modified during term reduction to reflect communication. The extra channel environment is provided to support possible aliasing of channels by treating channel variables as pointers or references into the channel environment. So, for a channel $c$ of session type $S$, a variable $x$ referring to this channel would be declared to have type $Chan~c$ and $c~:~S$ would be in the channel enviornment. The typing judgements and term reduction relation must be augmented to handle these environments. Unfortunately, a side-effect of this is that function types must specify the pre and post-states of channel environments and therefore, the type of arguments cannot work up to alpha-equivalence of channel identifiers. That is, if two channels $c$ and $d$ have the same session type, one cannot define a function to operate on variables of either channel type since the function definition will explicitly mention the identifier for the channel (in this case either, $Chan~c$ or $Chan~d$). However, the system does provide an attractive non-linear channel type which has been neglected in favour of a linear treatment of channels in more recent work~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2012}. Certainly, it is the case that an aliased approach would fit better with mainstream programming languages and perhaps the ``adoption and focus'' system~\cite{Fahndrich:2002} could offer some insights.

The work on a linear functional language with session types by \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the current project from the body of work on session-based $\lambda$-calculus type systems. They provide paper proofs of their systems type safety. Given the systems non-trivial features such as recursive session types and buffered channels for supporting asynchronous communication it will more difficult to extend these proofs manually. A mechanised version would provide stronger guarantees of correctness and allow others to more easily alter proofs as a result of changes made to the language.

@@

\citeauthor{Wadler:2012}'s recent work~\cite{Wadler:2012} brings together a process calculus, CP, based on the earlier work by \citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by \citeauthor{Gay:2010:LAST}. The work describes a continuation-passing style (CPS) translation from GV to CP treating session types as propositions in a classical linear logic. The process calculus is slightly different to that described in by $\pi$DILL~\cite{Caires:2010:STI}, apart from being classical in nature, the constructs for server replication are modified; CP provides a weakening and contraction rules whereas these are encoded in one rule for process reduction by \citeauthor{Caires:2010:STI}. Additionally, rather than a two-sided sequent calculus, \citeauthor{Wadler:2012} uses one-sided sequents for defining the process calculus. As presented, GV does not support all features of CP and thus the translation is one way (GV to CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to provide support for polymorphism and replication, following a similar CPS translation scheme.

@@

\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based type system providing session polymorphism. The system is more general than session polymorphism via subtyping~\cite{Gay:2005:SST}. The authors mechanise their type system and its properties in the Coq proof assistant. However, the system mechanised is restricted to the $\pi$-calculus and the main focus is proving soundness properties for the deduction rules allowing polymorphism on input, rather than a more general basis for studying programming with session types.

@@

>>>>

\usepackage{listings}
\usepackage{lstcoq}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{cmll}

\usepackage[style=alphabetic,sorting=nymt,natbib=true,
  maxbibnames=100]{biblatex}

\DeclareSortingScheme{nymt}{
  \sort{
    \field{author}
    \field{year}
    \field{month}
    \field{title}
  }
}

\addbibresource{proposal.bib}

@@

The aim of this project is to develop a framework for formalising lambda
calculus type systems involving session types using interactive theorem
proving (in particular, the system Coq). The motivation for the framework is
that it will enable a whole class of related languages and type systems to be
formalised with little extra proof effort. Inspired by previous work on a
mechanised framework for $\pi$-calculus type systems~\cite{Gay:2001:FFP}, I
shift to the lambda calculus to provide a basis on which session types for
mainstream programming languages can be studied, whilst taking advantage of
previous proof effort and specialising only for those parts unique to the
particular system. The framework will be heavily based on calculi developed by \citeauthor{Wadler:2012}~\cite{Wadler:2012} describing a relationship between classical linear logic (CP) and a functional language with session types (GV). GV is a minimal functional language supporting linearity so I believe such a basis will be general enough to support reasoning about a large class of functional languages with similar base properties.

From this framework, a number of directions can be explored including
session polymorphism session subtyping, and permitting different forms of
aliasing within the type system. The recent work by \citeauthor{Lindley:2014:SAP} \cite{Lindley:2014:SAP}, who extend GV to HGV supporting session forwarding, replication and polymorphism, would be an interesting place to start when looking to add these extensions.

%remove the adoption and focus stuff and add something about typestate

Additionally, the majority of previous session type systems for mainstream programming languages have focussed exclusively on linear type systems. One disadvantage of a linear type system is the need to re-bind objects after each operation~\cite{Gay:2010:LAST}. One could extend the GV type system to provide the benefits of aliasing with the guarantees of linearity which has been studied in several systems using a variety of alias control forms~\cite{Fahndrich:2002,Mazurak:2010:LLT,Tov:2011}. Another approach which has been utilised in previous session-based $\lambda$-calculus type systems, is to support side-effecting functions by allowing functions to close over channels in the environment~\cite{Gay:2003:STI,Vasconcelos:2006:TCM}. An interesting extension to the base GV type system would be to incorporate the closure semantics into the type system which would permit functions to modify session-typed channels in the environment. The typing rules would be similar to that described by \citeauthor{Gay:2010:MST} for typing expressions in an object-oriented setting for session types~\cite{Gay:2010:MST}, e.g. the rule for lambda abstraction could be:

... lambda abstraction ...

The $\Gamma$, $\Gamma '$ environments represent pre and post-states of variables satisfied by term $t$. The typing for the function argument is inspired by typestate-oriented programming~\cite{Aldrich:2009} and says that the function expects a channel behaving according to protocol $A$ and alters the state of the channel internally producing a channel behaving according to protocol $B$. It remains to be seen whether such typing rules can be encoded as described in the GV setting.

@@

most important: duality of sessions is obscured in $\pi$DILL because of the nature of the typing judgements

@@

Process reduction steps are equated with proof reduction via cut elimination. The relationship with proof normalisation in DILL provides absence of deadlocks in session-based communication for the process calculus.

@@

A series of works by \citeauthor{Gay:2003:STI}~\cite{Gay:2003:STI} and \citeauthor{Vasconcelos:2006:TCM}~\cite{Vasconcelos:2006:TCM} provide a $\lambda$-calculus based language with session types for channel primitives. The type system is nonstandard, featuring an additional environment for tracking the session types of channels; these types are modified during term reduction to reflect communication. The extra channel environment is provided to support possible aliasing of channels by treating channel variables as pointers or references into the channel environment. So, for a channel $c$ of session type $S$, a variable $x$ referring to this channel would be declared to have type $Chan~c$ and $c~:~S$ would be in the channel enviornment. The typing judgements and term reduction relation must be augmented to handle these environments. Unfortunately, a side-effect of this is that function types must specify the pre and post-states of channel environments and therefore, the type of arguments cannot work up to alpha-equivalence of channel identifiers. That is, if two channels $c$ and $d$ have the same session type, one cannot define a function to operate on variables of either channel type since the function definition will explicitly mention the identifier for the channel (in this case either, $Chan~c$ or $Chan~d$). However, the system does provide an attractive non-linear channel type which has been neglected in favour of a linear treatment of channels in more recent works~\cite{Gay:2010:LAST,Mazurak:2010:LCC,Wadler:2012}. Certainly, it is the case that an aliased approach would fit better with mainstream programming languages and perhaps studying a similar system could offer some insights.

The ``adoption and focus'' system~\cite{Fahndrich:2002} provides a notion of aliased access to linear resources. The type system requires special language features, e.g. to manage capabilities which can be thought as a form of object liveness mechanism, so GV typing judgements would need to be extended such that the capabilities can be included in soundness results. A extension to the translation to CP to incorporate the ``adopt'' and ``focus'' constructs would also need to be defined. It is not immediately clear how these features would assist in the management of session typed channels. The ``focus'' rule (and linearity generally) enforces an invariant on the focussed object (channel) such that the type is maintained during the focus operation:

... copy of focus operation here ...

Such a rule would appear to only benefit recursive session types. Indeed, focussing on a non-recursive channel would be pointless since one would be prevented from communicating across it since the focus rule requires maintenance of the channel type. Further, it would be undesirable to permit modification of a session type without guaranteeing exclusive access to the channel since this could lead to invalid communication e.g. trying to communicate on a closed channel.

The Alms language~\cite{Aldrich:2009} provides a functional language with affine types based on linear logic. A notion of type \textit{kind} is used to distinguish between affine and unrestricted values. The typing judgements use two contexts; one for variables with unrestricted kind and another for variables of affine type and type variables. Functions are also labelled with a kind which may depend on their arguments e.g. a multiple argument function must have affine kinding if it accepts an affine typed resource as an argument other than the final argument. The reason for this is simply to prevent the function from using an affine resource more than once. Additionally, functions act as closures over their environments so their qualifier may be determined by a free variable in the environment, and a relation is provided to calculate the appropriate kind qualifier. Type variables are provided to support kind polymorphism of function qualifiers. The system is expressive enough to implement session types as a module using a similar approach to \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} although treating channels as affine rather than linearly typed. \citeauthor{Aldrich:2009} note that Alms' type system can support the ``adoption and focus'' system but do not specify details or whether any changes are necessary to their typing judgements. Moreover, \citeauthor{Fahndrich:2002} do not provide operational semantics or soundness results for the ``adoption and focus'' type system.

The work on a linear functional language with session types by \citeauthor{Gay:2010:LAST}~\cite{Gay:2010:LAST} is the main influence for the current project from the body of work on session-based $\lambda$-calculus type systems. They provide paper proofs of their system's type safety. Given the non-trivial features of the type system such as recursive session types and buffered channels for supporting asynchronous communication it will be challenging to extend these proofs manually. A mechanised version would provide stronger guarantees of correctness and allow others to more easily alter proofs as a result of changes made to the language.

@@

\citeauthor{Wadler:2012}'s recent work~\cite{Wadler:2012} brings together a process calculus, CP, based on the earlier work by \citeauthor{Caires:2010:STI} with a functional language, GV, based on LAST by \citeauthor{Gay:2010:LAST}. The work describes a continuation-passing style (CPS) translation from GV to CP treating session types as propositions in a classical linear logic. The process calculus is slightly different to that described in by $\pi$DILL~\cite{Caires:2010:STI}, apart from being classical in nature, the constructs for server replication are modified; CP provides a weakening and contraction rules whereas these are encoded in one rule for process reduction by \citeauthor{Caires:2010:STI}. Additionally, rather than a two-sided sequent calculus, \citeauthor{Wadler:2012} uses one-sided sequents for defining the process calculus. As presented, GV does not support all features of CP and thus the translation is one way (GV to CP). \citeauthor{Lindley:2014:SAP}~\cite{Lindley:2014:SAP} extend GV to provide support for polymorphism and replication, following a similar CPS translation scheme. Lastly, \citeauthor{Lindley:2014:SPS}~\cite{Lindley:2014:SPS} describe operational semantics for GV and their system has greater similarities with LAST than prior GV works, for example, by treating send and receive as linear functions rather than language primitives.

@@

\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based type system providing session polymorphism. The system is more general than session polymorphism via subtyping~\cite{Gay:2005:SST}. more general how?

 The authors mechanise their type system and its properties in the Coq proof assistant. However, the system mechanised is restricted to the $\pi$-calculus and the main focus is proving soundness properties for the deduction rules allowing polymorphism on input, rather than a more general basis for studying programming with session types.

@@

I am going to look into adapting adoption and focus for the GV setting to allow one to control aliasing on session typed channels independent of their session characteristics so the aim is not to manipulate a linear type but to provide a nonlinear view of some resource and when changes are to be made, enforce some non-aliased view such that there no references in the context
perhaps the notion I'm looking for is to prevent sending on a guarded type or enforce the guarded type must be preserved in the adoption block
the trick in the adoption and focus system is that the type will be deallocated when its adopter's lifetime ends; I suspect in our system this is not useful since we will only be permitted to use a channel in a non-aliased context
